Canopy leaf area of a mature evergreen *Eucalyptus* woodland does not respond to elevated atmospheric [CO~2~] but tracks water availability

### 2015-08-19 manuscript.Rmd
Edit to reviewer's response; abstract minor edit; figure 4 thicker lines

Canopy leaf area, quantified by the leaf area index (LAI), is a crucial driver of forest productivity, water use, and energy balance. Because LAI responds to environmental drivers, it can represent an important feedback to climate change, but its responses to rising atmospheric [CO~2~] and water availability of forests have been poorly quantified. We studied canopy leaf area dynamics for `r nummon` months in a native evergreen *Eucalyptus* woodland exposed to Free Air CO~2~ Enrichment (the EucFACE experiment), in a subtropical climate where water limitation is common. We hypothesized that, because of expected stimulation of productivity and  water-use efficiency, LAI should increase with elevated [CO~2~]. We estimated LAI from diffuse canopy transmittance, and measured monthly leaf litter production. Contrary to expectation, LAI did not respond to elevated [CO~2~]. We found that LAI varied between `r f(rangeLAI[1])` and `r f(rangeLAI[2])` across the study period, but did not show a clear seasonal pattern. Instead, dynamics of LAI showed a quick increase after heavy rainfall, and a steady decrease during periods of low rainfall. Leaf litter production was correlated to changes in LAI, both during periods of decreasing LAI (when no leaf growth occurred), and during periods of increasing LAI (active shedding of old foliage when new leaf growth occurred). Leaf lifespan, estimated from mean LAI and total annual litter production, was up to two months longer under elevated [CO~2~] (`r f(LLtest$estimate2)` vs. `r f(LLtest$estimate1)` yrs; p `r pval(LLtest$p.value)`). Our main finding that LAI was not responsive to elevated CO~2~ is consistent with other forest FACE studies, but contrasts with the positive response of LAI commonly predicted by many ecosystem models.

>>>>

Canopy leaf area, quantified by the leaf area index (LAI), is a crucial driver of forest productivity, water use, and energy balance. Because LAI responds to environmental drivers, it can represent an important feedback to climate change, but its responses to rising atmospheric [CO~2~] and water availability of forests have been poorly quantified. We studied canopy leaf area dynamics for `r nummon` months in a native evergreen *Eucalyptus* woodland exposed to Free Air CO~2~ Enrichment (the EucFACE experiment), in a subtropical climate where water limitation is common. We hypothesized that, because of expected stimulation of productivity and  water-use efficiency, LAI should increase with elevated [CO~2~]. We estimated LAI from diffuse canopy transmittance, and measured monthly leaf litter production. Contrary to expectation, LAI did not respond to elevated [CO~2~]. We found that LAI varied between `r f(rangeLAI[1])` and `r f(rangeLAI[2])` across the study period. The dynamics of LAI showed a quick increase after heavy rainfall, and a steady decrease during periods of low rainfall. Leaf litter production was correlated to changes in LAI, both during periods of decreasing LAI (when no leaf growth occurred), and during periods of increasing LAI (active shedding of old foliage when new leaf growth occurred). Leaf lifespan, estimated from mean LAI and total annual litter production, was up to two months longer under elevated [CO~2~] (`r f(LLtest$estimate2)` vs. `r f(LLtest$estimate1)` yrs; p `r pval(LLtest$p.value)`). Our main finding that LAI was not responsive to elevated CO~2~ is consistent with other forest FACE studies, but contrasts with the positive response of LAI commonly predicted by many ecosystem models.

### 2015-08-19 review1.Rmd
Edit to reviewer's response; abstract minor edit; figure 4 thicker lines

---
output: word_document
---
In this manuscript, Duursma et al. present findings on the response of LAI and litter production to elevated atmospheric CO2 in EucFACE, a free air CO2 enrichment experiment established in a Eucalyptus woodland. The authors had expected elevated CO2 to increase LAI and litter production. In contrast to these expectations, the authors found CO2 to have no significant effect on either LAI or litter production over the 28-month experiment. However, they did find that CO2 extended leaf lifespan by approximately two months.

@@

**We now use pre-treatment BA as a covariate in a mixed-effects model. We found no CO2 or CO2 x time effect on LAI when BA was used as a covariate. There is also no effect of CO2 when BA is not included as a covariate, but due to the tight relationship between BA and LAI, using BA as a covariate increases the ability to find a CO2 effect. We only show the results when including BA as a covariate in the manuscript. Fig.2 now shows the LAI timeseries instead of the anomaly as before, and we have amended the text in various places to account for these changes.**

@@

**As noted above, we have redone the analysis and find no CO2 or CO2 x time effect. Although there was a brief period when the 95% confidence intervals were not overlapping (for the anomaly, though not when LAI is considered, see new Fig.2), this brief period did not result in a significant overall CO2 x time effect. For this reason we have not removed the insect attack from the study (it would have been appropriate if we did find a CO2 x time effect).**

@@

**The size, number, and relative coverage of the litterfall traps is broadly consistent with other forest FACE studies (see Table below), with the largest trap size. The size and coverage of the traps is therefore fully justified. The litter traps were placed in somewhat random locations across the ring, but never within XXX meters of the plot edge, to minimize contamination. We, as in all other FACE sites, cannot be completely certain that contamination did not occur, but it is likely that the vast majority of litter in each basket was produced in that FACE ring.**

@@

tabl <- tabl[,-ncol(tabl)]

@@

**We have moved this analysis, including the methods, to the Supporting Information. We believe this is interesting information for others concerned with LAI estimation, but agree it is a secondary objective only. We only briefly mention in the Methods that the two methods gave comparable results.**

@@

*TODO* **We have added a couple of sentences indicating the expectation of the litter response; it follows logically from the test of differences in LAI**

@@

**We have moved all methods-related content from the Results to the Methods section. We respond further to specific comments related to the Methods below**

@@

**Reworded to 'subtropical', which is more fitting** *IS THIS CORRECT*

@@

**todo**

@@

**FACE was defined in the 2nd paragraph of the definition, here the definition of EucFACE is quite obvious from the statement that the woodland is Eucalyptus-dominated**

@@

**Removed info on precipitation and PET here, as it is explained in more detail in the Methods already.**

@@

**If LAI changes, this must be due to changes in either the production of new foliage, or the lifespan of existing foliage. Both components can be assessed with measurements of litter production in combination with the time dynamics of LAI.** *Expand intro on litter fall a bit!!*

@@

**todo. which part is difficult???**

@@

**We have moved the discussion of discernable effect sizes to the discussion.**

@@

**Moved to methods. We disagree that this method 'assumes litter is grown and lost in the same time period'. Basically, gross leaf production must be equal to the net change in LAI (=the pool of leaves in the canopy) plus the amount of shedded leaf area, as measured by litter production. And to  "this calculation could indicate that a period of reduced Lt and large litter fall was associated with positive leaf production", this is only true if the reduction in LAI is less than the litter produced. In that case the canopy must have produced new leaves in that period to complete the sum of dLAI/dt = leafgrowth - litter. There are no assumptions on when the leaves were produced that show up in the litter traps at that time.**

**We think that the term 'litter production' may be misleading, as the term 'production' seems to indicate that the leaves were 'produced' during that time period, which we certainly do not assume nor imply (in fact, we have shown that leaf lifespan is ca. 1 year).**

@@

**It was not the purpose of the manuscript to generalize to other FACE sites, but we feel it is relevant discussion to the results of this study to the other FACE sites. This is informative because the results from the EucFACE do not fit well with a generalization across other FACE sites as presented by Norby and Zak (2011).**

@@

**We have moved Fig. 2 to the Supporting Information. We have retained the other figures as we feel they are essential to the manuscript (and six figures does not seem excessive to us).**

@@

**todo**

>>>>

---
output: word_document
---

**We thank the reviewer for the detailed and constructive comments. We have redone the data analysis to account for the reviewer's main concern, and provide detailed justification for the litter trap sampling design in the response below. We have accounted for the vast majority of detailed comments and suggestions, and have substantially edited the text for consistency, clarity and brevity.**

In this manuscript, Duursma et al. present findings on the response of LAI and litter production to elevated atmospheric CO2 in EucFACE, a free air CO2 enrichment experiment established in a Eucalyptus woodland. The authors had expected elevated CO2 to increase LAI and litter production. In contrast to these expectations, the authors found CO2 to have no significant effect on either LAI or litter production over the 28-month experiment. However, they did find that CO2 extended leaf lifespan by approximately two months.

@@

**We now use pre-treatment BA as a covariate in a mixed-effects model. We found no CO2 or CO2 x time effect on LAI when BA was used as a covariate. There is also no effect of CO2 when BA is not included as a covariate, but due to the tight relationship between BA and LAI, using BA as a covariate increases the ability to find a CO2 effect. On this point, we have also added a new figure in the supporting information (Fig. S5) that addresses the minimum detectable difference with a new method. Fig.2 now shows the LAI timeseries instead of the anomaly as before, and we have amended the text in various places to account for these changes.**

@@

**As noted above, we have redone the analysis and find no CO2 or CO2 x time effect. Although there was a brief period when the 95% confidence intervals were not overlapping (for the anomaly in the previous manuscript, though not when LAI is considered, see new Fig.2), this brief period did not result in a significant overall CO2 x time effect. For this reason we have not removed the insect attack from the study (it would have been appropriate if we did find a CO2 x time effect), since it does not confound the overall findings.**

@@

**The size, number, and relative coverage of the litterfall traps at EucFACE is consistent with the design of other forest FACE studies (see Table below), with in fact the largest trap size compared to other sites. The size and coverage of the traps is therefore fully justified. The litter traps were placed in somewhat random locations across the ring, but never within 2 meters of the plot edge, to minimize contamination. We, as in all other FACE sites, cannot be completely certain that contamination did not occur, but it is likely that the vast majority of litter in each basket was produced in that FACE ring.**

@@

tabl[,ncol(tabl)] <- 100*tabl[,ncol(tabl)]

@@

**We have moved this analysis, including the methods, in its entirety to the Supporting Information. We believe this is interesting information for others concerned with LAI estimation, but agree it is a secondary objective only. We only briefly mention in the Methods that the two methods gave comparable results.**

@@

**Since the reviewer deems the manuscript too lengthy already, we hesitate to include a full introduction of litter fall and our expectations. We have edited the final paragraph to make it more clear that we are also interested in the components of LAI dynamics, namely, leaf growth and litter production.**

@@

**We have moved all methods-related content from the Results to the Methods section, and substantially edited the entire section for brevity and clarity. We respond further to specific comments related to the Methods below.**

@@

**Reworded to 'subtropical', to be consistent with Methods.**

@@

**Removed from abstract.**

@@

**FACE was defined in the 2nd paragraph of the definition, here the definition of EucFACE is hopefully obvious from the statement that the woodland is Eucalyptus-dominated.**

@@

**Removed info on precipitation and PET here, as it is explained in more detail in the Methods already. In the Methods, we have indicated the period over which P and PET are estimated (both are longterm).**

@@

**If LAI changes over time between treatments, this must be due to changes in either the production of new foliage, or the lifespan of existing foliage. Both components can be assessed with measurements of litter production in combination with the time dynamics of LAI.**

@@

**I HAVE NOT DONE THIS - CAN SOMEONE TELL ME WHAT IS DIFFICULT ABOUT IT???**

@@

**We have moved the discussion of discernable effect sizes to the discussion, and the details of that analysis to the Supporting Information (new Fig. S5).**

@@

**Moved to methods (and edited for clarity). We disagree that this method 'assumes litter is grown and lost in the same time period'. Basically, gross leaf production must be equal to the net change in LAI (=the pool of leaves in the canopy) plus the amount of shedded leaf area, as measured by litter production. And to  "this calculation could indicate that a period of reduced Lt and large litter fall was associated with positive leaf production", this is only true if the reduction in LAI is less than the litter produced. In that case the canopy must have produced new leaves in that period to complete the sum of dLAI/dt = leafgrowth - litter. There are no assumptions on when the leaves were produced that show up in the litter traps at that time. We now explain this mass balance consideration in the Methods.**

@@

**It was not the purpose of the manuscript to generalize to other FACE sites, but we feel it is relevant to compare the results of this study to the other FACE sites. This is informative because the results from the EucFACE do not fit well with a generalization across other FACE sites as presented by Norby and Zak (2011).**

@@

**We have moved Fig. 2 to the Supporting Information. We have retained the other figures as we feel they are essential to the manuscript.**

@@
